### 2025-08-16 basic-statistics.qmd
20250816 - tweaks

There are many different types of scores and scales.
This book focuses on [raw scores](#sec-rawScores) and [*z*-scores](#sec-zScores).
For information on other scores and scales, including percentile ranks, *T*-scores, standard scores, scaled scores, and stanine scores, see here: <https://isaactpetersen.github.io/Principles-Psychological-Assessment/scoresScales.html#scoreTransformation> [@PetersenPrinciplesPsychAssessment].

@@

where $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), $\text{Cov}(x, y)$ is the covariance between $x$ and $y$, $s_x$ is the standard deviation of $x$, $s_y$ is the standard deviation of $y$, $n$ is the sample size, $z_x$ is the *z*-score for $x$, and $z_y$ is the *z*-score for $y$.

>>>>

There are many different types of scores and scales.
This book focuses on [raw scores](#sec-rawScores) and [*z* scores](#sec-zScores).
For information on other scores and scales, including percentile ranks, *T* scores, standard scores, scaled scores, and stanine scores, see here: <https://isaactpetersen.github.io/Principles-Psychological-Assessment/scoresScales.html#scoreTransformation> [@PetersenPrinciplesPsychAssessment].

@@

where $x$ is the [predictor variable](#sec-correlationalStudy), $y$ is the [outcome variable](#sec-correlationalStudy), $\text{Cov}(x, y)$ is the covariance between $x$ and $y$, $s_x$ is the standard deviation of $x$, $s_y$ is the standard deviation of $y$, $n$ is the sample size, $z_x$ is the [*z* scores](#sec-zScores) for $x$, and $z_y$ is the [*z* scores](#sec-zScores) for $y$.

### 2025-08-16 factor-analysis.qmd
20250816 - tweaks

An additional consideration is the scaling of the variables: whether to use raw variables, standardized variables, or dividing some variables by a constant to make the variables' variances more similar.
Before performing a [principal component analysis](#sec-pca) (PCA), it is generally important to ensure that the variables included in the PCA are on the same scale.
PCA seeks to identify components that explain variance in the data, so if the variables are not on the same scale, some variables may contribute considerably more variance than others.
A common way of ensuring that variables are on the same scale is to standardize them using, for example,  [*z*-scores](#zScores) that have a mean of zero and standard deviation of one.
By contrast, factor analysis can better accommodate variables that are on different scales.

>>>>

An additional consideration is the scaling of the variables: whether to use raw variables, standardized variables, or dividing some variables by a constant to make the variables' variances more similar.
Before performing a [principal component analysis](#sec-pca) (PCA), it is generally important to ensure that the variables included in the PCA are on the same scale.
PCA seeks to identify components that explain variance in the data, so if the variables are not on the same scale, some variables may contribute considerably more variance than others.
A common way of ensuring that variables are on the same scale is to standardize them using, for example, [*z* scores](#sec-zScores) that have a mean of zero and standard deviation of one.
By contrast, factor analysis can better accommodate variables that are on different scales.

### 2025-08-16 multiple-regression.qmd
20250816 - tweaks

Standardized regression coefficients can be obtained by standardizing the variables to [*z*-scores](#sec-zScores) so they all have a mean of zero and standard deviation of one.
The standardized regression coefficient ($\beta$) is interpreted such that, for every standard deviation change in the [predictor variable](#sec-correlationalStudy), there is a __ standard deviation change in the [outcome variable](#sec-correlationalStudy).
For instance, when examining the association between age and fantasy points, if the standardized regression coefficient is 0.1, players score on average 0.1 standard deviation more points for each additional standard deviation of their year of age.
Standardized regression coefficients—though not the case in all instances—tend to fall between [−1, 1].
Thus, standardized regression coefficients tend to be more comparable across variables and models compared to unstandardized regression coefficients.
In this way, standardized regression coefficients provide a meaningful index of [effect size](#sec-practicalSignificance).

@@

As noted in @sec-statisticalTestsPartitionedVariance, 
The coefficient of determination ($R^2$) reflects the proportion of variance in the [outcome (dependent) variable](#sec-correlationalStudy) that is explained by the model predictions, as in @eq-coefficientOfDeterminationPartitionedVariance: $R^2 = \frac{\text{variance explained in }Y}{\text{total variance in }Y}$.
Various formulas for $R^2$ are in @eq-rSquared.
Larger $R^2$ values indicate greater accuracy.
Multiple regression can be conceptualized with overlapping circles (similar to a venn diagram), where the non-overlapping portions of the circles reflect nonshared variance and the overlapping portions of the circles reflect shared variance, as in @fig-regression.

@@

  # total effect (c)
  total := abs(direct) + abs(indirect)
'
```

@@

  # total effect (c)
  total := abs(direct) + abs(indirectTotal)
'
```

>>>>

Standardized regression coefficients can be obtained by standardizing the variables to [*z* scores](#sec-zScores) so they all have a mean of zero and standard deviation of one.
The standardized regression coefficient ($\beta$) is interpreted such that, for every standard deviation change in the [predictor variable](#sec-correlationalStudy), there is a __ standard deviation change in the [outcome variable](#sec-correlationalStudy).
For instance, when examining the association between age and fantasy points, if the standardized regression coefficient is 0.1, players score on average 0.1 standard deviation more points for each additional standard deviation of their year of age.
Standardized regression coefficients—though not the case in all instances—tend to fall between [−1, 1].
Thus, standardized regression coefficients tend to be more comparable across variables and models compared to unstandardized regression coefficients.
In this way, standardized regression coefficients provide a meaningful index of [effect size](#sec-practicalSignificance) and can be used to identify the [predictors](#sec-correlationalStudy) with the strongest [predictive validity](#sec-predictiveValidity).

@@

As noted in @sec-statisticalTestsPartitionedVariance, 
The coefficient of determination ($R^2$) reflects the proportion of variance in the [outcome (dependent) variable](#sec-correlationalStudy) that is explained by the model predictions (i.e., by the [predictor variable(s)](#sec-correlationalStudy)), as in @eq-coefficientOfDeterminationPartitionedVariance: $R^2 = \frac{\text{variance explained in }Y}{\text{total variance in }Y}$.
Various formulas for $R^2$ are in @eq-rSquared.
Larger $R^2$ values indicate greater accuracy.
Multiple regression can be conceptualized with overlapping circles (similar to a venn diagram), where the non-overlapping portions of the circles reflect nonshared variance and the overlapping portions of the circles reflect shared variance, as in @fig-regression.

@@

  # total effect (c)
  total := abs(direct) + abs(indirect)

  # proportion mediated
  Pm := abs(indirect) / total
'
```

@@

  # total effect (c)
  total := abs(direct) + abs(indirectTotal)

  # proportion mediated
  Pm1 := abs(indirect1) / total
  Pm2 := abs(indirect2) / total
  PmTotal := abs(indirectTotal) / total
'
```
